# Week 1: SNS Study

## The Social Network Study

## Reading the data

R has several ways of reading data in. You data can be Raw plain files like CSV, tab delimited or specified by column width, for which you can use the [`readr`](https://cran.r-project.org/package=readr) package; or it can be binary files like dta (Stata), Octave, SPSS, for which [`foreign`](https://cran.r-project.org/package=readr) can be used; or it could be excel files in which case you should be using [`readxl`](https://cran.r-project.org/package=readxl). In our case, the data for this session is in Stata13 format, and so we will be using [`readstata13`](https://cran.r-project.org/package=readstata13).

```{r 03-read-data, message=FALSE}
library(dplyr)
library(magrittr)
library(readstata13)

# Reading the data
dat <- read.dta13("SNS datamerged081315edited.dta")

# Taking a look at the file
head(dat[,1:5])
```

For now, we won't be using all ~2,000 columns of this data (which is mostly because we have a wide format dataset here), so we need to filter all this data. In order to do so, we can use the `select` function from the `dplyr` package

```{r 03-filtering-data-pipe, cache=TRUE}
dat_filtered <- select(
  dat,
  School, photoid,
  matches("^sch_friend.+")
  )
```

The function `matches` allows us using regular expressions to select variables (reguwhat!?). Regular expressions are, as I once overheard, _the poor man's RA_. In this case, instead of us having to look for all the variables that start with the pattern `sch_friend`, we use regular expressions to catch that, character by character, we have:

*   `^`: Line start
*   `sch_friend`: Followed by `sch_friend`
*   `.+`: Followed by anything not null.

We will see more of this in the future. Now, if you don't want to use `dplyr` to do this simple variable selection, you can always use the base R function `subset`. The following example is equivalent to what we just did using `dplyr::select`, although it can be slower^[Besides of having nice syntax, `dplyr` offers optimized routines to handle your data. I personally use it specifically for that, as that can be nice when you are dealing with relatively large datasets.]


```{r 03-filter-no-pipe, eval=FALSE}
dat_filtered <- subset(
  dat,
  select = c(
    "School", "photoid",
    colnames(dat)[grepl("sch_friend.+", colnames(dat))]
    )
  )

```

As you can see, the `dplyr` syntax is more clear.

Now suppose that we want to create a unique id using the school and photo id. In this case, since both variables are numeric, a good way of doing it is to encode the id such that, for example, the last three `x` numbers are the photoid and the first ones are the school id. To do this we need to take into account the range of the variables. Here, `photoid` has the following range:

```{r 03-idrange}
(photo_id_ran <- range(dat_filtered$photoid))
```

As the variable spans up to `r photo_id_ran[2]`, we need to set the last `r nchar(photo_id_ran[2])` units of the variable to store the `photoid`. Again, we use `dplyr` to create this variable, and we will call it... `id` (mind blowing, right?):

```{r 03-newid}
(dat_filtered %<>% mutate(id = School*10000 + photoid)) %>%
  head %>%
  select(School, photoid, id)
```

Wow, what happend in the last three lines of code! What is that `%>%`? Well, that's the [piping operator](http://r4ds.had.co.nz/pipes.html), and it is a very nice way of writing nested function calls. In this case, instead of having write something like

```r
dat_filtered$id <- dat_filtered$School*10000 + dat_filtered$photoid
subset(head(dat_filtered), select = c(School, photoid, id))
```

```{r}
str(dat_filtered)
```




