# Week 1: SNS Study

## The Social Network Study

The data can be downloaded from

The codebook for the data provided here is in [the appendix](#sns-data).

## Today's Goals

1.  Read the data into R,

2.  Create a network with it,

3.  Compute descriptive statistics

4.  Visualize the network

## Data preprocessing

### Reading the data into R

R has several ways of reading data in. You data can be Raw plain files like CSV, tab delimited or specified by column width, for which you can use the [`readr`](https://cran.r-project.org/package=readr) package; or it can be binary files like dta (Stata), Octave, SPSS, for which [`foreign`](https://cran.r-project.org/package=readr) can be used; or it could be excel files in which case you should be using [`readxl`](https://cran.r-project.org/package=readxl). In our case, the data for this session is in Stata format:

```{r 03-read-data, message=FALSE}
library(dplyr)
library(magrittr)
library(foreign)

# Reading the data
dat <- foreign::read.dta("03-sns.dta")

# Taking a look at the data's first 5 columns and 5 rows
dat[1:5, 1:10]
```

### Creating a unique id for each participant

Now suppose that we want to create a unique id using the school and photo id. In this case, since both variables are numeric, a good way of doing it is to encode the id such that, for example, the last three `x` numbers are the photoid and the first ones are the school id. To do this we need to take into account the range of the variables. Here, `photoid` has the following range:

```{r 03-idrange}
(photo_id_ran <- range(dat$photoid))
```

As the variable spans up to `r photo_id_ran[2]`, we need to set the last `r nchar(photo_id_ran[2])` units of the variable to store the `photoid`. Again, we use `dplyr` to create this variable, and we will call it... `id` (mind blowing, right?):

```{r 03-newid}
(dat %<>% mutate(id = school*10000 + photoid)) %>%
  head %>%
  select(school, photoid, id)
```

Wow, what happend in the last three lines of code! What is that `%>%`? Well, that's the [piping operator](http://r4ds.had.co.nz/pipes.html), and it is a very nice way of writing nested function calls. In this case, instead of having write something like

```r
dat_filtered$id <- dat_filtered$school*10000 + dat_filtered$photoid
subset(head(dat_filtered), select = c(school, photoid, id))
```

## Creating a network

*   We want to build a social network. For that, we either use an adjacency matrix or an edgelist.

*   Each individual of the SNS data nomitated 19 friends from school. We will use those nominations to create the social network.

*   In this case, we will create the network by coercing the dataset into an edgelist.

### From survey to edgelist

Let's start by loading a couple of handy R packages for this task.

```{r 03-loading-tidyr-stringr, message=FALSE}
library(tidyr)
library(stringr)
```

Optionally, we can use the `tibble` type of object which is an alternative to the actual `data.frame`. This object is claimed to provide _more efficient methods for matrices and data frames_.

```{r 03-tibble}
dat <- as_tibble(dat)
```

What I like from tibbles is that when you print them on the console these actually look nice:
 
```{r 03-tibble-print}
dat
```



```{r 03-reshape}
# Maybe too much piping... but its cool!
net <- dat %>% 
  select(id, school, starts_with("sch_friend")) %>%
  gather(key = "varname", value = "content", -id, -school) %>%
  filter(!is.na(content)) %>%
  mutate(
    friendid = school*10000 + content,
    year     = str_extract(varname, "(?<=[a-z])[0-9]"),
    nnom     = str_extract(varname, "(?<=[a-z][0-9])[0-9]+")
  )
```

Let's take a look at this step by step:

1.  First, we subset the data: We want to keep `id, school, sch_friend*`. For the later we use the function `starts_with` (from the `tidyselect` package). This allows us to select all variables that starts with the word "`sch_friend`", which means that `sch_friend11, sch_friend12, ...` will all be selected.

    ```{r}
    dat %>% 
      select(id, school, starts_with("sch_friend"))
    ```
    
2.  Then, we reshape it to _long_ format: By transposing all the `sch_friend*` to long. We do this by means of the function `gather` (from the `tidyr` package). This is an alternative to the `reshape` function, and I personally find it easier to use. Let's see how it works:

    ```{r}
    dat %>% 
      select(id, school, starts_with("sch_friend")) %>%
      gather(key = "varname", value = "content", -id, -school)
    ```
    
    In this case the `key` parameter sets the name of the variable that will contain the name of the variable that was reshaped, while `value` is the name of the variable that will hold the content of the data (that's why I named those like that). The `-id, -school` bit tells the function to "drop" those variables before reshaping, in other words, "reshape everything but `id` and `school`".
    
    Also, notice that we passed from `r nrow(dat)` rows to 19 (nominations) * `r nrow(dat)` (subjects) * 4 (waves) = `r sprintf("%i", 19*nrow(dat)*4)` rows, as expected.
    
3.  As the nomination data can be empty for some cells, we need to take care of those cases, the `NA`s, so we filter the data:

    ```{r}
    dat %>% 
      select(id, school, starts_with("sch_friend")) %>%
      gather(key = "varname", value = "content", -id, -school) %>%
      filter(!is.na(content))
    ```
    
4.  And finally, we create three new variables from this dataset: `friendid`, `year`, and `nom_num` (nomination number). All this using regular expressions:
    
    ```{r}
    dat %>% 
      select(id, school, starts_with("sch_friend")) %>%
      gather(key = "varname", value = "content", -id, -school) %>%
      filter(!is.na(content)) %>%
      mutate(
        friendid = school*10000 + content,
        year     = str_extract(varname, "(?<=[a-z])[0-9]"),
        nnom     = str_extract(varname, "(?<=[a-z][0-9])[0-9]+")
        )
    ```
    
    The regular expression `(?<=[a-z])` matches a string that is preceeded by any letter from _a_ to _z_, whereas the expression `[0-9]` matches a single number. Hence, from the string `"sch_friend12"`, the regular expression will only match the `1`, as it is the only number followed by a letter. On the other hand, the expression `(?<=[a-z][0-9])` matches a string that is preceeded by a letter from _a_ to _z_ and a number from _0_ to _9_; and the expression `[0-9]+` matches a string of numbers--so it could be more than one. Hence, from the string `"sch_friend12"`, we will get `2`. We can actually se this 
    
    ```{r 03-miniregex}
    str_extract("sch_friend12", "(?<=[a-z])[0-9]")
    str_extract("sch_friend12", "(?<=[a-z][0-9])[0-9]+")
    ```
    
  
Now that we have this edgelist, we can create an igraph object

### igraph network

```{r 03-vertex}
vertices <- dat %>% 
  select(id, school, hispanic, female1, starts_with("eversmk"))
```

Let's now use the function `graph_from_data_frame` to create an `igraph` object:

```{r 03-igraph, message=FALSE, error = TRUE}
library(igraph)

ig_year1 <- net %>%
  filter(year == "1") %>%
  select(id, friendid) %>%
  graph_from_data_frame(
    vertices = vertices
    )
```

Ups! It seems that individuals are making nominations to other students that were not included on the survery. How to solve that? Well, it all depends on what you need to do! In this case, we will go for the _quietly-remove-em'-and-don't-tell_ strategy:

```{r 03-igraph-bis, cache=TRUE}

ig_year1 <- net %>%
  filter(year == "1") %>%
  
  # Extra line, all nominations must be in ego too.
  filter(friendid %in% id) %>% 
  
  select(id, friendid) %>%
  graph_from_data_frame(
    vertices = vertices
    )

ig_year1
```

So there we have, our network with `r vcount(ig_year1)` nodes and `r ecount(ig_year1)` edges. The next steps: get some descriptive stats and visualize our network.

## Network descriptive stats

While we could do all networks at once, in this part we will focus on computing some network statistics for one of the schools only. We start by school 111. The first question that you should be asking your self now is, "how can I get that information from the igraph object?." Well, vertex attributes and edges attributes can be accessed via the `V` and `E` functions respectively; moreover, we can list what vertex/edge attributes are available:

```{r 03-listing-attributes}
list.vertex.attributes(ig_year1)
list.edge.attributes(ig_year1) # we have no edge attributes here
```

Just like we would do with data frames, accessing vertex attributes is done via the dollar sign operator `$` together with the `V` function, for example, accessing the first 10 elements of the variable `hispanic` can be done as follows:

```{r 03-first-10-hispanic}
V(ig_year1)$hispanic[1:10]
```

Now that you know how to access vertex attributes, we can get the network corresponding to school 111 by identifying which vertices are part of it and pass that information to the `induced_subgraph` function:

```{r 03-igraph-year1-111}
# Which ids are from school 111?
school111ids <- which(V(ig_year1)$school == 111)

# Creating a subgraph
ig_year1_111 <- induced_subgraph(
  graph = ig_year1,
  vids  = school111ids
)
```

The `which` function in R returns a vector of indices indicating which elements are true. In our case it will return a vector of indices of the vertices which have the attribute `school` equal to 111. Now that we have our subgraph, we can compute different centrality measures^[For more information about the different centrality measurements, please take a look at the "Centrality" article on [Wikipedia](https://en.wikipedia.org/wiki/Centrality).] for each vertex and store them in the igraph object itself:

```{r 03-centrality-measures}
# Computing centrality measures for each vertex
V(ig_year1_111)$indegree   <- degree(ig_year1_111, mode = "in")
V(ig_year1_111)$outdegree  <- degree(ig_year1_111, mode = "out")
V(ig_year1_111)$closeness  <- closeness(ig_year1_111, mode = "total")
V(ig_year1_111)$betweeness <- betweenness(ig_year1_111, normalized = TRUE)
```

From here, we can _go back_ to our old habits and get the set of vertex attributes as a data frame so we can compute some summary statistics on the centrality measurements that we just got

```{r 03-centrality-stats}
# Extracting each vectex features as a data.frame
stats <- as_data_frame(ig_year1_111, what = "vertices")

# Computing quantiles for each variable
stats_degree <- with(stats, {
 cbind(
   indegree   = quantile(indegree, c(.025, .5, .975)),
   outdegree  = quantile(outdegree, c(.025, .5, .975)),
   closeness  = quantile(closeness, c(.025, .5, .975)),
   betweeness = quantile(betweeness, c(.025, .5, .975))
 )
})

stats_degree
```

The `with` function is somewhat similar to what `dplyr` allows us to do when we want to work with the dataset but without mentioning its name everytime that we ask for a variable. Without using the `with` function, the previous could have been done as follows:

```r
stats_degree <- 
 cbind(
   indegree   = quantile(stats$indegree, c(.025, .5, .975)),
   outdegree  = quantile(stats$outdegree, c(.025, .5, .975)),
   closeness  = quantile(stats$closeness, c(.025, .5, .975)),
   betweeness = quantile(stats$betweeness, c(.025, .5, .975))
 )

```

Now we will compute some statistics at the graph level:

```{r 03-network-stats}
cbind(
  size    = vcount(ig_year1_111),
  nedges  = ecount(ig_year1_111),
  density = edge_density(ig_year1_111),
  recip   = reciprocity(ig_year1_111),
  centr   = centr_betw(ig_year1_111)$centralization,
  pathLen = mean_distance(ig_year1_111)
  )
```



## Plotting the network in igraph

### Single plot

*   

Let's take a look at how does our network looks like when we use the default parameters in the plot method of the igraph object:

```{r 03-plot-raw, cache=TRUE, fig.cap='A not very nice network plot. This is what we get with the default parameters in igraph.', fig.align='center'}
plot(ig_year1)
```

Not very nice, right? A couple of things with this plot:

1.  We are looking at all schools simultaneously, which does not make sense.

2.  All the vertices have the same size, and more over, are overalaping.

3.  Given the number of vertices in these networks, the labels are not useful here.

```{r 03-color-hispanic}
col_hispanic <- V(ig_year1_111)$hispanic + 1
col_hispanic <- coalesce(col_hispanic, 3) 
col_hispanic <- c("steelblue", "tomato", "white")[col_hispanic]
```


```{r 03-plot-neat1, fig.cap="Friends network in time 1 for school 111. The graph excludes isolates."}
# Fancy graph
set.seed(1)
plot(
  ig_year1_111,
  vertex.size     = degree(ig_year1_111)/10 +1,
  vertex.label    = NA,
  edge.arrow.size = .25,
  layout          = layout_with_fr,
  vertex.color    = col_hispanic
  )
```

```{r 03-color-smoke}
col_eversmk1 <- V(ig_year1_111)$eversmk1 + 1
col_eversmk1 <- coalesce(col_eversmk1, 3) 
col_eversmk1 <- c("steelblue", "tomato", "white")[col_eversmk1]
```

```{r 03-plot-neat2, fig.cap="Friends network in time 1 for school 111. The graph excludes isolates."}
# Fancy graph
set.seed(1)
plot(
  ig_year1_111,
  vertex.size     = degree(ig_year1_111)/10 +1,
  vertex.label    = NA,
  edge.arrow.size = .25,
  layout          = layout_with_fr,
  vertex.color    = col_eversmk1
  )
```



We can actually write this as a function so that, instead of us copying and pasting the code $n$ times (supposing that we want to crate a plot similar to this $n$ times), we just need t

### Multiple plots

```{r 03-myplot-def}
myplot <- function(
  net,
  schoolid,
  mindgr = 1,
  vcol   = "tomato",
  ...) {
  
  # Creating a subgraph
  subnet <- induced_subgraph(
    net,
    which(degree(net, mode = "all") >= mindgr & V(net)$school == schoolid)
  )
  
  # Computing colors
  
  
  # Fancy graph
  set.seed(1)
  plot(
    subnet,
    vertex.size     = degree(subnet)/10,
    vertex.label    = NA,
    edge.arrow.size = .25,
    vertex.color    = vcol,
    layout          = layout_with_fr,
    ...
    )
}
```

```{r 03-myplot-call, fig.cap="All 5 schools in time 1. Again, the graphs exclude isolates."}
# Plotting all together
oldpar <- par(no.readonly = TRUE)
par(mfrow = c(2, 3), mai = rep(0, 4), oma= c(1, 0, 0, 0))
myplot(ig_year1, 111, vcol = "tomato")
myplot(ig_year1, 112, vcol = "steelblue")
myplot(ig_year1, 113, vcol = "black")
myplot(ig_year1, 114, vcol = "gold")
myplot(ig_year1, 115, vcol = "white")
par(oldpar)

# A fancy legend
legend(
  "bottomright",
  legend = c(111, 112, 113, 114, 115),
  pt.bg  = c("tomato", "steelblue", "black", "gold", "white"),
  pch    = 21,
  cex    = 1,
  bty    = "n",
  title  = "School"
  )
```

*   `oldpar <- par(no.readonly = TRUE)` This line stores the current parameters for plotting. Since we are going to be changing them, we better make sure we are able to go back!.

*   `par(mfrow = c(2, 3), mai = rep(0, 4), oma=rep(0, 4))` Here we are setting various things at the same time. `mfrow` specifies how many _figures_ will be drawn and in what order, in particular, we are asking the plotting device to allow for 2*3 = 6 plots organized in 2 rows and 3 columns, and these will be drawn by row.
    
    `mai` specifies the size of the margins in inches. Setting all margins equal to zero (which is what we are doing now) gives more space to the network itself. The same is true for `oma`. See `?par` for more info.
    

*   `myplot(ig_year1, ...)` This is simply calling our plotting function. The neat part of this is that, since we set `mfrow = c(2, 3)`, R takes care of _distributing_ the plots in the device.

*   `par(oldpar)` Finally, this line allows us to restore the plotting parameters.


    

